---
title: "Interactive Presentation with Plotly"
author: "Vitalii Diakonov"
date: "04/03/2024"
output:
  ioslides_presentation:
    widescreen: yes
    slide_level: 2
---

## Instructions

Create a web page presentation using R Markdown that features a plot created with Plotly. Host your webpage on either GitHub Pages, RPubs, or NeoCities. Your webpage must contain the date that you created the document, and it must contain a plot created with Plotly. We would love to see you show off your creativity!

## Review criteria

The rubric contains the following two questions:

1.  Does the web page feature a date and is this date less than two months before the date that you're grading this assignment?

2.  Is the web page a presentation and does it feature an interactive plot that appears to have been created with Plotly?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# Load necessary libraries
library(plotly)
```

## Plot 1: Sine Wave

```{r}
# Create sample data
x <- seq(0, 2*pi, length.out = 100)
y <- sin(x)

# Create a plot with Plotly
plot1 <- plot_ly(x = ~x, y = ~y, type = 'scatter', mode = 'lines')

# Customize layout
layout <- list(title = "Sine Wave", xaxis = list(title = "X-axis"), yaxis = list(title = "Y-axis"))

# Display the plot
plot1 <- plot1 %>% layout(layout)
plot1

```

## Plot 2: Bar Chart

```{r}
# Create sample data
data <- data.frame(
  category = c("A", "B", "C", "D"),
  value = c(10, 20, 15, 25)
)

# Create a bar chart with Plotly
plot2 <- plot_ly(data, x = ~category, y = ~value, type = 'bar')

# Customize layout
layout <- list(title = "Bar Chart", xaxis = list(title = "Category"), yaxis = list(title = "Value"))

# Display the plot
plot2 <- plot2 %>% layout(layout)
plot2
```

## Plot 3: Scatter plot

```{r}
# Create sample data
x <- rnorm(100)
y <- rnorm(100)

# Create a scatter plot with Plotly
plot3 <- plot_ly(x = ~x, y = ~y, type = 'scatter', mode = 'markers')

# Customize layout
layout <- list(title = "Scatter Plot", xaxis = list(title = "X-axis"), yaxis = list(title = "Y-axis"))

# Display the plot
plot3 <- plot3 %>% layout(layout)
plot3
```

## Conclusion

Thank you for viewing my interactive presentation featuring Plotly plots. I hope you enjoyed it!
