---
title: "Reproducible Pitch Presentation"
author: "Vitalii Diakonov"
date: "2024-04-04"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Gapminder Data Exploration

-   **Presenter**: Vitalii Diakonov
-   **Date**: 04/03/2024

This presentation introduces a Shiny application designed to explore various global health and economic indicators provided by the Gapminder dataset.


## Slide 2: The Gapminder Dataset

-   **Global Data**: Life Expectancy, GDP per Capita, Population.
-   **Insights**: Understand global trends over decades.
-   **Application**: Visualize and analyze country-specific and global patterns.

The Gapminder dataset provides a unique lens through which we can observe the development and health of nations over time.


## Slide 3: Shiny Application Features

-   **Interactive Exploration**: Users can select countries, continents, and time ranges to explore.
-   **Dynamic Visualizations**: Life expectancy and GDP per capita over time.
-   **User-Friendly Interface**: Designed for ease of use without sacrificing depth of analysis.

The application aims to make data exploration accessible and informative.

## Slide 4: Demonstrating the Application

```{r example}
# Example R code simulating part of the application's functionality
library(ggplot2)
library(gapminder)
data(gapminder)

ggplot(gapminder[gapminder$country %in% c("China", "United States") & gapminder$year > 1952 & gapminder$year < 2007,], aes(x = year, y = lifeExp, color = country)) +
  geom_line() + geom_point() +
  labs(title = "Life Expectancy Over Time", 
       subtitle = "Data from Gapminder for selected years",
       x = "Year", y = "Life Expectancy") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```


## Slide 5: Conclusion and Further Information

-   **Goal**: Make data exploration engaging and accessible.
-   **Learn More**: Visit [Gapminder](https://www.gapminder.org/data/) for more information on the dataset.
-   **Try the App**: [Lifeapp](https://psydoc19.shinyapps.io/lifeapp/).

Thank you for your attention! For questions or further information, feel free to reach out.
