---
title: "Interactive Map with Leaflet"
author: "Vitalii Diakonov"
date: "04/03/2024"
output: html_document
---

## Instructions

Create a web page using R Markdown that features a map created with Leaflet. Host your webpage on either GitHub Pages, RPubs, or NeoCities.

Your webpage must contain the date that you created the document, and it must contain a map created with Leaflet. We would love to see you show off your creativity!

## Review criteria

The rubric contains the following two questions:

1.  Does the web page feature a date and is this date less than two months before the date that you're grading this assignment?

2.  Does the web page feature an interactive map that appears to have been created with Leaflet?

## **Interactive Map of Top 10 Beaches in Florida**

This map showcases the top 10 beaches in Florida, each marked with a custom icon. Hover over or click on a marker to read a brief description of each beach.

```{r setup, include=FALSE}
# Load necessary libraries
library(leaflet)
library(magrittr) # for the pipe operator
```

```{r map-creation, echo=FALSE}
# Beaches data: names, coordinates, and descriptions
beaches <- data.frame(
  name = c("Clearwater Beach", "Siesta Key Beach", "South Beach", "Sanibel Island Beach", "Destin Beach", 
           "Fort Lauderdale Beach", "Naples Beach", "Pensacola Beach", "Key West Beaches", "St. Pete Beach"),
  lat = c(27.9778, 27.2670, 25.7617, 26.4483, 30.3935, 26.1224, 26.1420, 30.3328, 24.5551, 27.7253),
  lng = c(-82.8276, -82.5463, -80.1918, -82.0225, -86.4958, -80.1043, -81.7948, -87.1498, -81.7800, -82.7412),
  description = c("Popular for its white sands and clear waters.", "Famous for its quartz sand and vibrant sunsets.",
                  "Iconic beach with lively nightlife.", "Tranquil island with unique seashells.",
                  "Emerald waters and white sands.", "Beautiful urban beaches and luxury hotels.",
                  "Upscale beach with fine dining and shopping.", "Home to the longest pier in the Gulf of Mexico.",
                  "Laid-back atmosphere with historic sites.", "Award-winning beach with soft, sugary sands.")
)

# Define a custom icon for beach markers
beachIcon <- icons(
  iconUrl = "https://raw.githubusercontent.com/pointhi/leaflet-color-markers/master/img/marker-icon-2x-yellow.png",
  iconWidth = 25, iconHeight = 41,
  iconAnchorX = 12, iconAnchorY = 41
)

# Create the Leaflet map
leaflet(beaches) %>% 
  addProviderTiles(providers$Esri.WorldImagery) %>% # Use Esri World Imagery for the base map
  addMarkers(~lng, ~lat, popup = ~paste("<b>", name, "</b><br>", description), icon = beachIcon) %>% # Add custom icon markers with popups
  addLayersControl(
    baseGroups = c("Esri World Imagery"), # Add layer control to toggle base map
    options = layersControlOptions(collapsed = FALSE)
  )
```
